---
title: "Settlement Severity Index"
subtitle: "Sensitivity Analysis"
output:
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: true
    theme: readable
    fig_width: 8
    fig_height: 6
  word_document:
    toc: no
    toc_depth: '2'
    output_dir: "Outputs"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "GEU R&D"
params:
  admin_level_index: NULL
  selected_index_method: NULL
  data_file: NULL
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, eval = T, include = F, warning = F, message = F, cache.lazy = T)
```

```{r Packages}
gc()
```

```{r Parameters}
input_path <- "input/"

admin_boundaries <- c("admin1", "admin2", "admin3", "admin4")
```

```{r Import data}
# params$data_file is always set when rendered from Shiny (via rmarkdown::render).
# Only fall back to list.files() when genuinely knitting locally without params
# (i.e. params$data_file is NULL and we are NOT inside a Shiny process).
if (!is.null(params$data_file)) {
  
  # --- Shiny / render() path -------------------------------------------
  data_index <- readRDS(params$data_file) %>%
    mutate(severity_value = as.numeric(severity_value))
  
} else {
  
  # --- Local knit-only fallback (development) ---------------------------
  local_file <- list.files(pattern = "severity_index_data_hti", full.names = TRUE)
  
  if (length(local_file) == 0) {
    stop(
      "No data file found. When rendering from Shiny, pass params$data_file. ",
      "When knitting locally, make sure a file matching 'severity_index_data_hti' ",
      "exists in the working directory."
    )
  }
  
  data_index <- read.xlsx(local_file[1], sheet = "Raw data") %>%
    mutate(severity_value = as.numeric(severity_value))
  
}

# Keep only admin boundary columns that actually exist in the data
admin_boundaries <- admin_boundaries[admin_boundaries %in% names(data_index)]

# Settlement-level data is only meaningful when admin4 is present.
# Without admin4, the lowest granularity is admin3 (area level).
has_settlement <- "admin4" %in% admin_boundaries
```

```{r Detect available index columns}
# ── Detect which flag/proportion/score columns exist ──────────────────────────
# The selected_index_method param drives which columns we highlight, but all
# plots gracefully fall back to whatever columns are actually present.

# Method passed in from Shiny (or NULL when knitting locally)
selected_method <- if (!is.null(params$selected_index_method)) {
  params$selected_index_method
} else {
  # Auto-detect a sensible default from what exists
  if ("flag4" %in% names(data_index))        "flag4"
  else if ("flag3" %in% names(data_index))   "flag3"
  else if ("flag4_plus" %in% names(data_index)) "flag4+"
  else if ("proportion4_settlement" %in% names(data_index)) "proportion4"
  else if ("score_index_settlement" %in% names(data_index)) "score"
  else NULL
}

# Helper: map a method string to its column name fragments
# Returns a named list: flag_col, proportion_settlement_col, proportion_sector_col,
#                       settlement_col, sector_col, area_col, label
resolve_method_cols <- function(method, df_names) {
  
  candidates <- list(
    "flag3" = list(
      flag_col              = "flag3",
      proportion_settlement = "proportion3_settlement",
      proportion_sector     = "proportion3_sector",
      settlement_col        = "flag3_settlement",
      sector_col            = "flag3_sector",
      area_col              = "mean_flag3_area",
      label                 = "Flag 3"
    ),
    "flag4" = list(
      flag_col              = "flag4",
      proportion_settlement = "proportion4_settlement",
      proportion_sector     = "proportion4_sector",
      settlement_col        = "flag4_settlement",
      sector_col            = "flag4_sector",
      area_col              = "mean_flag4_area",
      label                 = "Flag 4"
    ),
    "flag4+" = list(
      flag_col              = "flag4_plus",
      proportion_settlement = "proportion4_plus_settlement",
      proportion_sector     = "proportion4_plus_sector",
      settlement_col        = "flag4_plus_settlement",
      sector_col            = "flag4_plus_sector",
      area_col              = "mean_flag4_plus_area",
      label                 = "Flag 4+"
    ),
    "proportion3" = list(
      flag_col              = "flag3",
      proportion_settlement = "proportion3_settlement",
      proportion_sector     = "proportion3_sector",
      settlement_col        = "proportion3_settlement",
      sector_col            = "proportion3_sector",
      area_col              = "mean_proportion3_area",
      label                 = "Proportion 3"
    ),
    "proportion4" = list(
      flag_col              = "flag4",
      proportion_settlement = "proportion4_settlement",
      proportion_sector     = "proportion4_sector",
      settlement_col        = "proportion4_settlement",
      sector_col            = "proportion4_sector",
      area_col              = "mean_proportion4_area",
      label                 = "Proportion 4"
    ),
    "proportion4+" = list(
      flag_col              = "flag4_plus",
      proportion_settlement = "proportion4_plus_settlement",
      proportion_sector     = "proportion4_plus_sector",
      settlement_col        = "proportion4_plus_settlement",
      sector_col            = "proportion4_plus_sector",
      area_col              = "mean_proportion4_plus_area",
      label                 = "Proportion 4+"
    ),
    "score" = list(
      flag_col              = NULL,
      proportion_settlement = "proportion4_settlement",  # fallback for freq section
      proportion_sector     = "proportion4_sector",
      settlement_col        = "score_index_settlement",
      sector_col            = "score_index_sector",
      area_col              = "score_index_area",
      label                 = "Score Index"
    )
  )
  
  # Resolve the requested method; fall back gracefully
  cols <- if (length(method) == 1L && !is.na(method) && method %in% names(candidates)) {
    candidates[[method]]
  } else {
    # Pick the first method whose flag_col actually exists in the data
    found <- NULL
    for (m in names(candidates)) {
      fc <- candidates[[m]]$flag_col
      if (is.null(fc) || (length(fc) == 1L && fc %in% df_names)) {
        found <- candidates[[m]]
        break
      }
    }
    if (is.null(found)) candidates[[1]] else found
  }

  # Null-out any column name that doesn't actually exist in the data.
  # "label" is a plain string, not a column name, so it must be preserved.
  col_entries <- setdiff(names(cols), "label")
  cols[col_entries] <- lapply(cols[col_entries], function(v) {
    if (is.null(v) || !is.character(v) || length(v) != 1L) return(v)
    if (v %in% df_names) v else NULL
  })
  cols
}

method_cols <- resolve_method_cols(selected_method, names(data_index))

# Convenience accessor with a clear error message
col_exists <- function(col) !is.null(col) && col %in% names(data_index)

# Sector colour palette (extend freely)
custom_colors <- c(
  "fsl" = "#F08080",
  "shl" = "#D2B48C",
  "wsh" = "#ADD8E6",
  "hea" = "#DDA0DD",
  "pop" = "#FFD700",
  "pro" = "#A9A9A9",
  "cm"  = "#F4A460",
  "edu" = "#90EE90"
)

# Sectors actually present in the data
available_sectors <- sort(unique(data_index$sector))
```

# Sensitivity Analysis — `r method_cols$label` {.tabset}

```{r method-summary, include=TRUE, results='asis'}
active_label <- if (!is.null(method_cols$label) && nchar(method_cols$label) > 0) {
  method_cols$label
} else if (!is.null(selected_method)) {
  selected_method   # fall back to the raw param string
} else {
  "Unknown (no method selected)"
}

cat("> **Active method:**", active_label, "\n\n")
cat("> **Admin levels used:**", paste(admin_boundaries, collapse = " › "), "\n\n")
cat("> **Settlement-level analysis:**",
    ifelse(has_settlement,
           "✔ Available (admin4 present)",
           "✘ Not available — lowest granularity is area (admin3)"),
    "\n\n")
```

---

## Missing Values

Missing values matter because the Flag Index counts indicators that
*exceed* a severity threshold. Values that are missing are treated as
"not severe", so it is important to confirm that assumption is valid.

```{r flag_na_data}
data_na <- data_index %>%
  filter(is.na(severity_value))
```

```{r flag_na_plot, include=TRUE}
if (nrow(data_na) == 0) {
  ggplot() +
    annotate("text", x = 1, y = 1,
             label = "No NA values in this Index :)",
             size = 6, hjust = 0.5, vjust = 0.5) +
    theme_void()
} else {
  ggplot(data_na, aes(x = question)) +
    geom_bar() +
    coord_flip() +
    labs(title = "Distribution of NAs Across Questions",
         x = "Question", y = "Number of NAs") +
    theme_minimal()
}
```

---

## Severity Frequency

How often is each indicator above the active severity threshold? A very
high frequency may indicate a threshold that is too lenient — or genuine
widespread severity. Contextual knowledge is essential here.

```{r severity_freq_plot, include=TRUE}
flag_col      <- method_cols$flag_col
freq_y_label  <- if (has_settlement) "Number of Settlements" else "Number of Areas (admin3)"

if (col_exists(flag_col)) {
  
  data_flagged <- data_index %>%
    filter(.data[[flag_col]] == 1) %>%
    group_by(across(all_of(admin_boundaries)), question, sector) %>%
    summarise(count = n(), .groups = "drop")
  
  if (nrow(data_flagged) == 0) {
    ggplot() +
      annotate("text", x = 1, y = 1,
               label = paste0("No indicators flagged above ", method_cols$label, " threshold."),
               size = 5, hjust = 0.5, vjust = 0.5) +
      theme_void()
  } else {
    ggplot(data_flagged,
           aes(x = reorder(question, count, FUN = sum), y = count, fill = sector)) +
      geom_bar(stat = "identity") +
      coord_flip() +
      labs(
        title = paste("Indicators above", method_cols$label, "Severity"),
        x = "Indicator", y = freq_y_label
      ) +
      scale_fill_manual(values = custom_colors, na.value = "grey70") +
      theme_minimal()
  }
  
} else {
  # Score method — show mean score per indicator instead
  if ("score_index_sector" %in% names(data_index)) {
    data_score_summary <- data_index %>%
      group_by(question, sector) %>%
      summarise(mean_score = mean(score_index_sector, na.rm = TRUE), .groups = "drop")
    
    ggplot(data_score_summary,
           aes(x = reorder(question, mean_score), y = mean_score, fill = sector)) +
      geom_bar(stat = "identity") +
      coord_flip() +
      labs(title = "Mean Score Index per Indicator",
           x = "Indicator", y = "Mean Score") +
      scale_fill_manual(values = custom_colors, na.value = "grey70") +
      theme_minimal()
  } else {
    ggplot() +
      annotate("text", x = 1, y = 1,
               label = paste("Column", flag_col, "not found in data."),
               size = 5, hjust = 0.5) +
      theme_void()
  }
}
```

---

## Distribution

```{r distribution_header, include=TRUE, results='asis'}
if (has_settlement) {
  cat("Distribution of the selected index across **settlements** (admin4).\n\n")
} else {
  cat(
    "> ⚠️ **No settlement-level data:** `admin4` is not present in this dataset.",
    "The distribution below is shown at the **area level** (admin3) instead.\n\n"
  )
}
```

```{r distribution_plot, include=TRUE}
# Choose the appropriate column and grouping level depending on data granularity
if (has_settlement) {
  dist_col   <- method_cols$settlement_col
  dist_level <- "Settlements"
  # Unique rows at settlement level
  data_dist <- data_index %>%
    select(all_of(admin_boundaries),
           any_of(names(data_index)[grepl("settlement|score_index_settlement", names(data_index))])) %>%
    distinct()
} else {
  dist_col   <- method_cols$area_col
  dist_level <- "Areas (admin3)"
  # Unique rows at area level (admin1–admin3)
  area_admins <- admin_boundaries[admin_boundaries != "admin4"]
  data_dist <- data_index %>%
    select(all_of(area_admins),
           any_of(names(data_index)[grepl("area|score_index_area", names(data_index))])) %>%
    distinct()
}

if (col_exists(dist_col)) {

  x_label <- dplyr::case_when(
    grepl("proportion", dist_col) ~ paste("Proportion of indicators —", method_cols$label),
    grepl("score",      dist_col) ~ paste("Score Index (", tolower(dist_level), ")", sep = ""),
    grepl("flag|mean",  dist_col) ~ paste("# indicators flagged —", method_cols$label),
    TRUE ~ dist_col
  )

  ggplot(data_dist, aes(x = .data[[dist_col]])) +
    geom_bar() +
    labs(
      title = paste("Distribution of", method_cols$label, "Index across", dist_level),
      x = x_label,
      y = paste("Number of", dist_level)
    ) +
    theme_minimal()

} else {
  ggplot() +
    annotate("text", x = 1, y = 1,
             label = paste0("Column '", dist_col, "' not available in data."),
             size = 5, hjust = 0.5) +
    theme_void()
}
```

---

## Sector Information

Maximum observed index value per sector for the active method.

```{r sector_max_data}
sector_col <- method_cols$sector_col

if (col_exists(sector_col)) {
  max_values <- data_index %>%
    group_by(sector) %>%
    summarise(max_value = max(.data[[sector_col]], na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(max_value))
} else {
  max_values <- NULL
}
```

```{r sector_max_plot, include=TRUE}
if (!is.null(max_values) && nrow(max_values) > 0) {
  ggplot(max_values, aes(x = reorder(sector, max_value), y = max_value)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    coord_flip() +
    labs(
      title = paste("Maximum", method_cols$label, "Index per Sector"),
      x = "Sector",
      y = paste("Max", method_cols$label)
    ) +
    theme_minimal()
} else {
  ggplot() +
    annotate("text", x = 1, y = 1,
             label = paste("Sector column '", sector_col, "' not available in data.", sep = ""),
             size = 5, hjust = 0.5) +
    theme_void()
}
```

*These bars represent the maximum possible severity per sector under the
selected method, which is useful when considering normalisation.*

---

## Distributions & Correlations {.tabset}

Severity value distributions and inter-indicator correlation matrices,
split by sector. Only sectors present in the data are shown.

```{r sector_tabs, include=TRUE, results='asis'}

for (sec in available_sectors) {
  
  # Open tab
  cat("\n\n###", toupper(sec), "\n\n")
  
  data_sector <- data_index %>% filter(sector == sec)
  
  # ── Ridge plot ───────────────────────────────────────────────────────────────
  if (nrow(data_sector) > 0 && length(unique(data_sector$question)) >= 1) {
    
    p_ridge <- ggplot(data_sector,
                      aes(x = severity_value, y = question, fill = question)) +
      geom_density_ridges(scale = 0.9) +
      theme_ridges() +
      theme(legend.position = "none") +
      scale_x_continuous(limits = c(1, 5), breaks = 1:5) +
      labs(title = paste("Severity Value Distribution —", toupper(sec)),
           x = "Severity Value", y = NULL)
    
    print(p_ridge)
    cat("\n\n")
    
  } else {
    cat("*No data available for sector:", sec, "*\n\n")
    next
  }
  
  # ── Correlation matrix ───────────────────────────────────────────────────────
  data_sector_wide <- data_sector %>%
    select(all_of(admin_boundaries), question, severity_value) %>%
    pivot_wider(names_from = question, values_from = severity_value) %>%
    select(-all_of(admin_boundaries))
  
  # Need at least 2 columns and 2 complete rows for a meaningful correlation
  n_cols <- ncol(data_sector_wide)
  n_complete <- sum(complete.cases(data_sector_wide))
  
  if (n_cols >= 2 && n_complete >= 2) {
    sector_corr <- tryCatch(
      cor(data_sector_wide, use = "complete.obs"),
      error = function(e) NULL
    )
    
    if (!is.null(sector_corr) && !all(is.na(sector_corr))) {
      corrplot(sector_corr,
               method   = "color",
               type     = "upper",
               tl.col   = "black",
               tl.srt   = 45,
               title    = paste("Correlation Matrix —", toupper(sec)),
               mar      = c(0, 0, 2, 0))
    } else {
      cat("*Correlation could not be computed for sector:", sec, "*\n\n")
    }
  } else if (n_cols == 1) {
    cat("*Only one indicator in this sector — correlation matrix not applicable.*\n\n")
  } else {
    cat("*Not enough complete observations to compute correlation matrix.*\n\n")
  }
  
  cat("\n\n")
}
```
