---
title: "Severity Index Report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
params:
  admin_level_index: NULL
  selected_index_method: NULL
  select_admin_bounds: NULL
  data_file: NULL
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(
  echo       = FALSE,
  eval       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  results    = "asis",
  fig.width  = 10,
  fig.height = 5
)
```

```{r load-data, include=FALSE, results="hide"}
if (!is.null(params$data_file)) {
  data <- readRDS(params$data_file)
} else {
  local_file <- list.files(pattern = "index_data\\.rds", full.names = TRUE, recursive = TRUE)
  if (length(local_file) == 0) stop("No local index_data.rds found.")
  data <- readRDS(local_file[1])
}
data <- data %>% mutate(severity_value = as.numeric(severity_value))
```

```{r resolve-params, include=FALSE, results="hide"}

# ---- Admin boundaries -------------------------------------------------------
all_possible_admins <- c("admin1", "admin2", "admin3", "admin4")

select_admin_bounds <- if (!is.null(params$select_admin_bounds)) {
  params$select_admin_bounds
} else {
  all_possible_admins[all_possible_admins %in% names(data)]
}

admin_boundaries <- all_possible_admins[
  all_possible_admins %in% select_admin_bounds &
  all_possible_admins %in% names(data)
]

has_settlement <- "admin4" %in% admin_boundaries
lowest_admin   <- tail(admin_boundaries, 1)

# User-selected area level from the Shiny dropdown (e.g. "admin2")
admin_level <- if (!is.null(params$admin_level_index) && length(params$admin_level_index) >= 1) {
  al <- params$admin_level_index[[1]]
  if (al %in% admin_boundaries) al else head(admin_boundaries, 1)
} else {
  head(admin_boundaries, 1)
}

# Admin columns that are *strictly above* the lowest level (used for area re-aggregation)
area_admins <- admin_boundaries[seq_len(which(admin_boundaries == admin_level))]

# ---- Methods ----------------------------------------------------------------
# Normalise whatever Shiny passes (key or full label) to the short key used
# in resolve_method_cols. Shiny selectInput should pass the choice VALUE not
# the label, but we guard against both just in case.
normalise_method <- function(m) {
  lookup <- c(
    "Flag Severity 3 (Indicator, Cluster, Settlement, Area)"    = "flag3",
    "Flag Severity 4 (Indicator, Cluster, Settlement, Area)"    = "flag4",
    "Flag Severity 4+ (Indicator, Cluster, Settlement, Area)"   = "flag4+",
    "Proportion Severity 3 (Cluster, Settlement, Area)"         = "proportion3",
    "Proportion Severity 4 (Cluster, Settlement, Area)"         = "proportion4",
    "Proportion Severity 4+ (Cluster, Settlement, Area)"        = "proportion4+",
    "Score Index (Indicator, Cluster, Settlement, Area)"        = "score"
  )
  if (m %in% names(lookup)) lookup[[m]] else m
}

selected_methods <- if (!is.null(params$selected_index_method) &&
                        length(params$selected_index_method) > 0) {
  sapply(params$selected_index_method, normalise_method, USE.NAMES = FALSE)
} else {
  if      ("flag4"                  %in% names(data)) "flag4"
  else if ("flag3"                  %in% names(data)) "flag3"
  else if ("score_index_settlement" %in% names(data)) "score"
  else stop("Could not detect any supported index method in the data.")
}

# ---- Column resolver --------------------------------------------------------
#
# Key design decisions baked in here:
#
# indicator_col:
#   - flag methods      -> the binary flag column itself (flag3 / flag4 / flag4_plus)
#                          Summing it across rows gives "how many units flagged this indicator"
#   - proportion methods -> ALSO the corresponding flag column.
#                          proportion4 and proportion4+ differ by their flag threshold,
#                          so using flag4 vs flag4_plus gives distinct indicator rankings.
#                          (proportion_not_recommended is identical for all proportion
#                           methods because it is computed before thresholding.)
#   - score             -> severity_value (raw 1-5 per indicator x unit)
#
# area_col:
#   - Stored at the LOWEST admin grain (admin3 or admin4).
#   - When the user selects a coarser area level (e.g. admin2), we re-aggregate
#     by taking the mean across all lowest-level units within that area.
#     This is done in the area tab, NOT in the resolver.
#
resolve_method_cols <- function(method, df_names) {
  candidates <- list(
    "flag3" = list(
      flag_col       = "flag3",
      indicator_col  = "flag3",
      settlement_col = "flag3_settlement",
      sector_col     = "flag3_sector",
      area_col       = "mean_flag3_area",
      label          = "Flag 3"
    ),
    "flag4" = list(
      flag_col       = "flag4",
      indicator_col  = "flag4",
      settlement_col = "flag4_settlement",
      sector_col     = "flag4_sector",
      area_col       = "mean_flag4_area",
      label          = "Flag 4"
    ),
    "flag4+" = list(
      flag_col       = "flag4_plus",
      indicator_col  = "flag4_plus",
      settlement_col = "flag4_plus_settlement",
      sector_col     = "flag4_plus_sector",
      area_col       = "mean_flag4_plus_area",
      label          = "Flag 4+"
    ),
    "proportion3" = list(
      flag_col       = "flag3",
      indicator_col  = "flag3",           # use flag to distinguish from proportion4
      settlement_col = "proportion3_settlement",
      sector_col     = "proportion3_sector",
      area_col       = "mean_proportion3_area",
      label          = "Proportion 3"
    ),
    "proportion4" = list(
      flag_col       = "flag4",
      indicator_col  = "flag4",           # distinct from proportion4+ (flag4 vs flag4_plus)
      settlement_col = "proportion4_settlement",
      sector_col     = "proportion4_sector",
      area_col       = "mean_proportion4_area",
      label          = "Proportion 4"
    ),
    "proportion4+" = list(
      flag_col       = "flag4_plus",
      indicator_col  = "flag4_plus",      # distinct from proportion4
      settlement_col = "proportion4_plus_settlement",
      sector_col     = "proportion4_plus_sector",
      area_col       = "mean_proportion4_plus_area",
      label          = "Proportion 4+"
    ),
    "score" = list(
      flag_col       = NULL,
      indicator_col  = "severity_value",
      settlement_col = "score_index_settlement",
      sector_col     = "score_index_sector",
      area_col       = "score_index_area",
      label          = "Score Index"
    )
  )

  # Exact match first; then auto-detect by first method whose flag_col exists
  cols <- if (length(method) == 1L && !is.na(method) && method %in% names(candidates)) {
    candidates[[method]]
  } else {
    found <- NULL
    for (m in names(candidates)) {
      fc <- candidates[[m]]$flag_col
      if (is.null(fc) || (length(fc) == 1L && fc %in% df_names)) { found <- candidates[[m]]; break }
    }
    if (is.null(found)) candidates[[1]] else found
  }

  # Null-out entries whose column does not exist in the data (preserve label)
  col_entries <- setdiff(names(cols), "label")
  cols[col_entries] <- lapply(cols[col_entries], function(v) {
    if (is.null(v) || !is.character(v) || length(v) != 1L) return(v)
    if (v %in% df_names) v else NULL
  })
  cols
}

col_exists <- function(col) !is.null(col) && length(col) == 1L && col %in% names(data)

# ---- Colour palette ---------------------------------------------------------
custom_colors <- c(
  "fsl" = "#F08080", "shl" = "#D2B48C", "shelter" = "#D2B48C",
  "wsh" = "#ADD8E6", "wash" = "#ADD8E6", "hea" = "#DDA0DD",
  "pop" = "#FFD700", "pro" = "#A9A9A9", "cm"  = "#F4A460",
  "edu" = "#90EE90", "aap" = "#FFB6C1"
)

# ---- Shared helpers ---------------------------------------------------------
styled_kable <- function(df) {
  df %>%
    knitr::kable(format = "html", escape = FALSE) %>%
    kableExtra::kable_styling(
      full_width        = FALSE,
      bootstrap_options = c("hover", "condensed", "responsive")
    )
}

lollipop_chart <- function(df, x_col, y_col, title = "", y_label = "") {
  ggplot(df, aes(x = reorder(.data[[x_col]], .data[[y_col]]),
                 y = .data[[y_col]])) +
    geom_segment(aes(xend = .data[[x_col]], yend = 0),
                 colour = "grey70", linewidth = 0.8) +
    geom_point(colour = "#2C7BB6", size = 4) +
    coord_flip() + theme_minimal() +
    labs(title = title, x = "", y = y_label)
}

bar_chart <- function(df, x_col, y_col, fill_col = NULL, title = "", y_label = "") {
  p <- ggplot(df, aes(x = reorder(.data[[x_col]], .data[[y_col]]),
                      y = .data[[y_col]])) +
    geom_bar(stat = "identity",
             aes(fill = if (!is.null(fill_col)) .data[[fill_col]] else NULL)) +
    coord_flip() + theme_minimal() +
    labs(title = title, x = "", y = y_label, fill = "Sector")
  if (!is.null(fill_col)) p <- p + scale_fill_manual(values = custom_colors, na.value = "grey70")
  p
}

# Safely re-aggregate area_col from lowest admin level up to a coarser admin level.
# area_col values are stored at lowest_admin grain; when admin_level is coarser
# (e.g. admin2 when data is at admin3), we take the mean across children.
aggregate_to_admin_level <- function(df, area_col, target_admins, all_admins) {
  df %>%
    select(all_of(all_admins), all_of(area_col)) %>%
    distinct() %>%
    group_by(across(all_of(target_admins))) %>%
    summarise(value = mean(.data[[area_col]], na.rm = TRUE), .groups = "drop")
}
```

---

```{r report-header, results='asis'}
active_labels <- sapply(selected_methods, function(m) resolve_method_cols(m, names(data))$label)
cat("## Report Summary\n\n")
cat(paste0(
  "| | |\n|---|---|\n",
  "| **Method(s)** | ",            paste(active_labels,    collapse = ", "), " |\n",
  "| **Admin levels in data** | ", paste(admin_boundaries, collapse = " > "), " |\n",
  "| **Finest unit tab** | ",      lowest_admin, " |\n",
  "| **Selected area tab** | ",    admin_level,  " |\n",
  "| **Settlement data** | ",
  ifelse(has_settlement, "Available (admin4 present)",
         paste0("Not available - finest grain is ", lowest_admin)), " |\n\n"
))
```

---

# Results {.tabset}

## Indicator {.tabset}

For **flag/proportion** methods: total number of lowest-level units where each indicator is flagged (strictly method-specific — flag4 and flag4+ will differ because they use different thresholds).
For **score**: mean severity value (1-5) per indicator across all units.

```{r indicator-tabs, results='asis'}
for (method in selected_methods) {

  mc        <- resolve_method_cols(method, names(data))
  ind_col   <- mc$indicator_col   # flag col for flag/proportion; severity_value for score
  is_score  <- method == "score"
  label     <- mc$label

  cat(paste0("\n### ", label, "\n\n"))

  if (!col_exists(ind_col)) {
    cat(paste0("*Column `", ind_col, "` not found for method `", method, "`.*\n\n")); next
  }

  if (is_score) {
    # Mean severity value per indicator (1-5 scale)
    top_ind <- data %>%
      select(question, sector, severity_value) %>%
      group_by(question, sector) %>%
      summarise(value = mean(severity_value, na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(value)) %>%
      slice_head(n = 10) %>%
      mutate(Rank = row_number(), .before = 1, value = round(value, 2))

    y_label <- "Mean severity value (1-5)"

  } else {
    # flag / proportion: count how many rows (units x indicator) are flagged.
    # For proportion methods indicator_col = flag_col so proportion4 uses flag4
    # and proportion4+ uses flag4_plus -> they produce distinct counts.
    top_ind <- data %>%
      select(question, sector, all_of(ind_col)) %>%
      group_by(question, sector) %>%
      summarise(
        n_flagged = sum(.data[[ind_col]], na.rm = TRUE),
        .groups   = "drop"
      ) %>%
      arrange(desc(n_flagged)) %>%
      slice_head(n = 10) %>%
      mutate(Rank = row_number(), .before = 1)

    # Also compute % of total units flagged for this indicator
    n_units_total <- data %>% select(all_of(admin_boundaries)) %>% distinct() %>% nrow()
    top_ind <- top_ind %>%
      mutate(pct = round(100 * n_flagged / n_units_total, 1))

    y_label <- paste("Units flagged (%) -", label)

    top_ind <- top_ind %>% rename(value = pct)
  }

  top_ind_display <- top_ind %>%
    rename(Indicator = question, Sector = sector, !!y_label := value)
  if ("n_flagged" %in% names(top_ind_display)) {
    top_ind_display <- top_ind_display %>% rename(`Units flagged (n)` = n_flagged)
  }

  top_ind_display %>% styled_kable() %>% cat()
  cat("\n\n")

  print(bar_chart(top_ind, "question", "value", fill_col = "sector",
                  title = paste("Top Indicators -", label), y_label = y_label))
  cat("\n\n")
}
```

## Sector {.tabset}

For **flag** methods: % of area-level units where the sector has at least one flagged indicator (denominator = total area-units).
For **proportion** methods: mean % of indicators flagged within each sector, averaged across area-units (NAs excluded where indicator count < 50% threshold).
For **score**: mean score index (25th-pct interpolation) per sector, averaged across area-units.

```{r sector-tabs, results='asis'}
# Sector columns are computed at area×sector grain (evaluated_admin_bounds in the server,
# which corresponds to area_admins in the report). We must deduplicate at that grain,
# NOT at the full admin_boundaries grain, to avoid double-counting across settlements.
#
# Column semantics (from si_functions2.R):
#
#   flag methods      -> flagX_sector  = SUM of flagX per area×sector  (integer >= 0)
#                        Interpretation: how many indicators in that sector were flagged
#                        in this area-unit. To get "% of area-units with any flag in
#                        sector", check > 0 and divide by total n area-units.
#
#   proportion methods -> proportionX_sector = % of indicators flagged in that area×sector
#                        (already a proportion, 0-100). Display as mean across area-units.
#
#   score             -> score_index_sector = 25th-pct score per area×sector (continuous).
#                        Display as mean across area-units.

for (method in selected_methods) {

  mc         <- resolve_method_cols(method, names(data))
  sector_col <- mc$sector_col
  label      <- mc$label
  is_flag    <- method %in% c("flag3", "flag4", "flag4+")
  is_prop    <- str_detect(method, "proportion")
  is_score   <- method == "score"

  cat(paste0("\n### ", label, "\n\n"))

  if (!col_exists(sector_col)) {
    cat(paste0("*Column `", sector_col, "` not available for method `", method, "`.*\n\n")); next
  }

  # Deduplicate at area×sector grain.
  # area_admins are the admin columns up to and including the user-selected admin_level.
  # Sector columns from the server are computed at evaluated_admin_bounds = area_admins.
  sec_data <- data %>%
    select(all_of(area_admins), sector, all_of(sector_col)) %>%
    distinct()

  # n_area_units: number of distinct area-level units (the correct denominator for flag %)
  n_area_units <- data %>% select(all_of(area_admins)) %>% distinct() %>% nrow()

  if (is_flag) {
    # flagX_sector = count of flagged indicators in this area×sector.
    # Derive: was the sector flagged at all in this area-unit? (> 0)
    # Then report: % of area-units where this sector had >= 1 flag.
    top_sec <- sec_data %>%
      group_by(sector) %>%
      summarise(
        n_flagged   = sum(.data[[sector_col]] > 0, na.rm = TRUE),
        pct_flagged = round(100 * n_flagged / n_area_units, 1),
        .groups = "drop"
      ) %>%
      arrange(desc(pct_flagged)) %>%
      mutate(Rank = row_number(), .before = 1)

    y_label <- paste("% area-units with sector flagged -", label)

    top_sec %>%
      mutate(fraction = paste0(n_flagged, " / ", n_area_units)) %>%
      rename(Sector = sector,
             `Area-units with flag (n / total)` = fraction,
             !!y_label := pct_flagged) %>%
      select(-n_flagged) %>%
      styled_kable() %>% cat()
    cat("\n\n")

    print(bar_chart(top_sec, "sector", "pct_flagged", fill_col = "sector",
                    title = paste("Sector -", label), y_label = y_label))
    cat("\n\n")

  } else if (is_prop) {
    # proportionX_sector = % of indicators in this sector that are flagged, per area×sector.
    # Display as the mean proportion across all area-units (mirrors how area_col is shown).
    # Also surface error_message_sector where applicable.
    has_err <- "error_message_sector" %in% names(data)

    err_df <- if (has_err) {
      data %>%
        select(all_of(area_admins), sector, error_message_sector) %>%
        distinct() %>%
        group_by(sector) %>%
        summarise(error_message = first(na.omit(error_message_sector)), .groups = "drop")
    } else NULL

    top_sec <- sec_data %>%
      group_by(sector) %>%
      summarise(value = mean(.data[[sector_col]], na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(value)) %>%
      mutate(Rank = row_number(), .before = 1)

    if (!is.null(err_df)) {
      top_sec <- left_join(top_sec, err_df, by = "sector")
    } else {
      top_sec$error_message <- NA_character_
    }

    top_sec <- top_sec %>%
      mutate(display = ifelse(is.na(value),
                              coalesce(error_message, "N/A"),
                              paste0(round(value, 1), "%")))

    y_label <- paste("Mean % indicators flagged per area-unit -", label)

    top_sec %>%
      select(Rank, Sector = sector, !!y_label := display) %>%
      styled_kable() %>% cat()
    cat("\n\n")

    chart_data <- top_sec %>% filter(!is.na(value))
    if (nrow(chart_data) > 0) {
      print(bar_chart(chart_data, "sector", "value", fill_col = "sector",
                      title = paste("Sector -", label), y_label = y_label))
    }
    cat("\n\n")

    # Note on sectors below threshold
    if (has_err) {
      err_sectors <- data %>%
        select(sector, num_avail_ind_sector, total_ind_per_sector, error_message_sector) %>%
        distinct() %>%
        filter(!is.na(error_message_sector) & error_message_sector != "") %>%
        group_by(sector) %>%
        summarise(
          avail  = first(na.omit(num_avail_ind_sector)),
          total  = first(na.omit(total_ind_per_sector)),
          reason = first(na.omit(error_message_sector)),
          .groups = "drop"
        )
      if (nrow(err_sectors) > 0) {
        cat(paste0(
          "> **Note — sectors excluded or NAs present:** The following sectors had ",
          "insufficient available indicators (< 50% of required) in at least one area-unit. ",
          "Means above exclude those NA area-units.\n\n"
        ))
        err_sectors %>%
          mutate(`Available / Required` = paste0(avail, " / ", total)) %>%
          rename(Sector = sector, Reason = reason) %>%
          select(Sector, `Available / Required`, Reason) %>%
          styled_kable() %>% cat()
        cat("\n\n")
      }
    }

  } else {
    # Score: score_index_sector = 25th-pct interpolated score per area×sector.
    # Display as mean across area-units.
    has_err <- "error_message_sector" %in% names(data)

    top_sec <- sec_data %>%
      group_by(sector) %>%
      summarise(value = mean(.data[[sector_col]], na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(value)) %>%
      mutate(Rank = row_number(), .before = 1)

    if (has_err) {
      err_df <- data %>%
        select(sector, error_message_sector) %>%
        distinct() %>%
        group_by(sector) %>%
        summarise(error_message = first(na.omit(error_message_sector)), .groups = "drop")
      top_sec <- left_join(top_sec, err_df, by = "sector")
    } else {
      top_sec$error_message <- NA_character_
    }

    top_sec <- top_sec %>%
      mutate(display = ifelse(is.na(value),
                              coalesce(error_message, "N/A"),
                              as.character(round(value, 2))))

    y_label <- "Mean score index per sector"

    top_sec %>%
      select(Rank, Sector = sector, !!y_label := display) %>%
      styled_kable() %>% cat()
    cat("\n\n")

    chart_data <- top_sec %>% filter(!is.na(value))
    if (nrow(chart_data) > 0) {
      print(bar_chart(chart_data, "sector", "value", fill_col = "sector",
                      title = paste("Sector -", label), y_label = y_label))
    }
    cat("\n\n")
  }
}
```

## `r paste0("Lowest unit (", lowest_admin, ")")` {.tabset}

Rankings at the finest available administrative level: **`r lowest_admin`**.
`r if (!has_settlement) paste0("> No admin4 column present - ", lowest_admin, " is the finest grain available.")`

```{r lowest-unit-tabs, results='asis'}
for (method in selected_methods) {

  mc    <- resolve_method_cols(method, names(data))
  label <- mc$label

  # Use settlement col when admin4 exists; fall back to area col otherwise
  rank_col <- if (has_settlement && col_exists(mc$settlement_col)) {
    mc$settlement_col
  } else if (col_exists(mc$area_col)) {
    mc$area_col
  } else {
    NULL
  }

  cat(paste0("\n### ", label, "\n\n"))

  if (is.null(rank_col)) {
    cat(paste0("*No column found for method `", method, "` at unit level.*\n\n")); next
  }

  # Group by all admin_boundaries and take first() of rank_col to guarantee
  # exactly one row per lowest-level unit (deduplicates across indicator rows
  # even if rank_col somehow varies within a unit due to data structure).
  top_units <- data %>%
    select(all_of(admin_boundaries), all_of(rank_col)) %>%
    group_by(across(all_of(admin_boundaries))) %>%
    summarise(across(all_of(rank_col), ~ first(na.omit(.x))), .groups = "drop") %>%
    arrange(desc(.data[[rank_col]])) %>%
    slice_head(n = 10) %>%
    mutate(Rank = row_number(), .before = 1)

  val_label <- paste(lowest_admin, "value -", label)

  top_units %>%
    rename(!!val_label := all_of(rank_col)) %>%
    styled_kable() %>% cat()
  cat("\n\n")

  print(lollipop_chart(top_units, lowest_admin, rank_col,
                       title = paste("Top", lowest_admin, "units -", label),
                       y_label = val_label))
  cat("\n\n")
}
```

## `r paste0("Area (", admin_level, ")")` {.tabset}

Rankings at the user-selected area level: **`r admin_level`**.

`r if (admin_level == lowest_admin) paste0("> Selected level (", admin_level, ") is the same as the finest unit - results match the tab above.")` 

The area value is the **mean** of all `r lowest_admin`-level values within each `r admin_level`.

```{r area-tabs, results='asis'}
for (method in selected_methods) {

  mc       <- resolve_method_cols(method, names(data))
  area_col <- mc$area_col
  label    <- mc$label

  cat(paste0("\n### ", label, "\n\n"))

  if (!col_exists(area_col)) {
    cat(paste0("*Column `", area_col, "` not found for method `", method, "`.*\n\n")); next
  }

  # area_col is stored at the LOWEST admin grain (e.g. admin3).
  # Step 1: deduplicate to one row per admin3 (group_by all admins, take first).
  # Step 2: if admin_level is coarser than lowest_admin, re-aggregate upward
  #         by taking the mean across children within each admin_level unit.
  unit_level <- data %>%
    select(all_of(admin_boundaries), all_of(area_col)) %>%
    group_by(across(all_of(admin_boundaries))) %>%
    summarise(area_value = first(na.omit(.data[[area_col]])), .groups = "drop")

  top_areas <- unit_level %>%
    group_by(across(all_of(area_admins))) %>%
    summarise(value = mean(area_value, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(value)) %>%
    slice_head(n = 10) %>%
    mutate(Rank = row_number(), .before = 1)

  area_label <- paste(admin_level, "mean value -", label)

  top_areas %>%
    rename(!!area_label := value) %>%
    styled_kable() %>% cat()
  cat("\n\n")

  print(lollipop_chart(top_areas, admin_level, "value",
                       title = paste("Top areas (", admin_level, ") -", label),
                       y_label = area_label))
  cat("\n\n")
}
```
